### 2025-10-30 Paper/paperdraft_Sep2024.tex
Updates from Overleaf

    \begin{abstract}
		{\small
         Managers are often seen as drivers of firm decisions, strategies, and performance. Further, manager characteristics have been linked to employee wages, decisions of mergers and acquisitions, and financial performance. Health care policies in the US often focus on improving quality of care without increasing costs, and yet it is unclear whether hospital managers might contribute to this goal. In this paper, I estimate the differential response to such quality incentives by different types of nonprofit hospital executives: those with and without a clinical training background. I find that hospitals without clinically trained executives respond more to the financial incentives than those with clinically trained executives. I provide evidence that this may be due to differing preferences for profit vs. patient welfare. Thus, incentivizing clinically trained executives in hospitals could be an effective method to improve value in the US health care system. 
		} 
	\end{abstract}

>>>>

    \begin{abstract}
		{\small
         Managers are often seen as drivers of firm decisions, strategies, and performance. In addition, characteristics of managers have been related to employee wages, mergers and acquisition decisions, and financial performance. Health care policies in the US often focus on improving quality of care without increasing costs, but it is unclear how different types of leaders might interact with this goal. In this paper, I estimate the differential response to these quality incentives by different types of nonprofit hospital executives: those with and without a clinical training background. Using a synthetic difference-in-differences methodology, I show that nonclinical executive teams internalize financial incentives on quality more than clinical executive teams. Further, nonclinical executive teams respond to the incentives more similarly to for-profit hospitals, indicating that this may be due to differing preferences for profit versus patient welfare. 
		} 
	\end{abstract}

### 2025-10-30 Paper/paperdraft_fall2025.tex
Updates from Overleaf

	\author{\vspace{10mm}Hanna Glenn\footnote{Department of Economics, Emory University, 1602 Fishburne Drive, Atlanta, GA 30322, hanna.glenn@emory.edu.} }

@@

    \begin{abstract}
		{\small
         Managers are often seen as drivers of firm decisions, strategies, and performance. Further, manager characteristics have been linked to employee wages, decisions of mergers and acquisitions, and financial performance. Health care policies in the US often focus on improving quality of care without increasing costs, and yet it is unclear whether hospital managers might contribute to this goal. In this paper, I estimate the differential response to such quality incentives by different types of nonprofit hospital executives: those with and without a clinical training background. I find that hospitals without clinically trained executives respond more to the financial incentives than those with clinically trained executives. I provide evidence that this may be due to differing preferences for profit vs. patient welfare. Thus, incentivizing clinically trained executives in hospitals could be an effective method to improve value in the US health care system. 
		} 
	\end{abstract}

@@

    In corporate environments, executives are considered crucial drivers of a company's practices and performance, as they are heavily involved in the culture, visions, and strategies of the firm. Executive heterogeneity has been linked to various actions and outcomes of firms; characteristics such as gender, age, and background are correlated with financial performance and behaviors such as employee diversity, wages, aggressiveness in mergers and acquisitions, and other corporate strategies.\footnote{\cite{bertrand2003managing}; \cite{matsa2011chipping}; \cite{matsa2011chipping}; \cite{adams2009women}; \cite{ahern2012changing}; \cite{flabbi2019female}; \cite{benmelech2015military}; \cite{custodio2013generalists}; \cite{frydman2019rising}} However, one aspect of firm leadership that is not well understood is how different leaders may interact with changing policy incentives differently.

    In this paper, I study whether type of executive training affects firm behavior in the context of US hospitals. In the US, health care makes up almost 20\% of spending (\cite{CMSNationalHealthExpenditure}), and a major goal of policymakers is to improve the quality of health care without increasing costs. Yet, little is known about how hospital managers affect health care value, and it is unclear whether existing findings on executive characteristics can be extrapolated to the health care, and particularly hospital, context. Existing literature is generally limited to large, publicly traded firms, whereas a majority of US hospitals are nonprofit with presumably different objectives and policy incentives.

    A characteristic relevant to this context is whether or not hospital executives have clinical training. Physicians can bring a unique combination of clinical and administration expertise, and they have insights that can potentially benefit their employees and lead to better quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives is controversial, as physicians are not always trained in business management, which can lead to poor practices and lower quality (\cite{HarvardBusinessReview2018}). Thus, I estimate how clinically trained executives in hospital leadership affects hospital outcomes. I scrape executive names and titles of nonprofit hospitals in the US from publicly available Tax Form 990s. Combining this with other public hospital data, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofits due to data availability, this sample of hospitals is an important driver of health care in the US. Private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up 36\% of hospitals and staff 107 beds on average (\cite{ASPE_2023}).

@@

    I investigate whether the result might be explained by differing preferences on the trade-off between profit and patient welfare, where I model these hospital objectives and allow weight placed on each to vary with executive team type. Hospitals that place more weight on profit will naturally respond more drastically to a policy that adds quality directly into the profit function. Thus, a potential reason for the difference in response is that non-clinical executive teams are more profit driven. I investigate this empirically by comparing the response of hospitals with and without clinically trained executives to the response of for-profit hospitals, who have revealed that they are more profit driven by their ownership type. I find that nonprofit hospitals without clinically trained executives respond to the financial incentives on quality similarly to for-profits, while nonprofits without clinically trained executives respond differently than for-profits. For-profits and non-clinical executive teams change quality more in response to the incentive change than clinically trained executive teams.

    Finally, the question remains whether clinically trained executives are a signal of unchanging underlying hospital preferences, or if they actively manage the hospital differently, effectively changing hospital objectives. I analyze this by leveraging changes in hospital propensity to hire clinical executives to decompose the full effect of clinical leadership into a signaling effect (clinical executives signal underlying hospital preferences), or managing effect (clinical executives manage the hospital differently). First, I show that hospital executive changes are not correlated with the change to financial incentives. For the signaling effect, I compare hospitals who ever have a clinically trained executive to hospitals who never have a clinically trained executive. For the managing effect, I limit to hospitals who ever have a clinically trained executive and compare those who have a clinically trained executive when the policies were announced to those who have clinically trained executives at other times. I find that the difference in response is driven by clinically trained executives managing the hospital differently.

@@

    \section{Theoretical Framework}

@@

    \noindent where $\pi(\theta)$ is a profit function depending on quality $\theta$, $u(.)$ is extra utility gained from quality, which is increasing and concave in $\theta$, and $\alpha\in[0,1]$ captures the weight placed on profit vs. societal benefit. There is an implicit stay-open condition throughout the model.

    In this paper, I consider hospital choice of quality under pay-for-performance incentives, which essentially incorporate quality directly into a hospital's profit function (\cite{dranove2011health}). Alternatively, a fee-for-service system only incentivizes quantity. A hospital that places all value on profit will only care about quality when it enters into the profit function, and therefore will respond more drastically to pay-for-performance than a hospital who already cared about quality through societal benefit before the incentives. I will now formalize this intuition mathematically. For simplicity, I abstract away from quality having a direct impact on quantity of patients or price.\footnote{As long as the relationship between quality and quantity/prove are not dependent on clinical experience on the leadership team, this does not change the conclusion of the model.} Taking the first order condition yields:

@@

    Now, consider incentives when the hospital is in a strictly fee-for-service environment. That is, quality benefits patients, but hospitals do not directly derive profit from having a high quality. Then, assuming there is a positive cost to increasing quality, $\pi(\theta)$ is decreasing in $\theta$. Then $\frac{du'(\theta)}{d\alpha} > 0$. Using the Implicit Function, I can write $\frac{d\theta}{d\alpha}$ as 
    $\frac{du'(\theta)/d\alpha}{du'(\theta)/d\theta}<0$, revealing that in a fee-for-service setting, the more weight the hospital places on profit, the lower quality of care chosen.

    Next, assume the hospital is in a strictly pay-for-performance setting, where quality benefits patients and directly increases profit. Then, assuming the marginal financial benefit of increasing quality is greater than the marginal financial cost, $\pi(\theta)$ is increasing in $\theta$. Then $\frac{du'(\theta)}{d\alpha} < 0$, revealing that in a pay-for-performance setting, the more weight placed on profit, the higher quality of care chosen by the hospital. I combine the results found in each scenario into one response function that depends on $\alpha$, where $\theta_{p4p}$ is quality under pay-for-performance incentives and $\theta_{ffs}$ is quality under fee-for-service only. That is,

@@

    Hence, this model predicts that change in quality depends on how much weight the hospital places on profit vs. societal benefit. Particularly, hospitals with more weight on profit respond more than hospitals with more weight placed on societal benefit. Therefore, for leadership characteristics that are plausibly correlated with how much weight is placed on patient welfare versus quality, such as clinical training, we expect to see a differential response to pay-for-performance incentives being enacted.

@@

    For each hospital in the tax form data, I merge to hospital matches in the AHA data based on name and location.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.} For all hospitals that can be matched to a general hospital in the AHA data, I use optimal character recognition (OCR) methods to extract the names, titles, and positions of each listed executive from its Tax Form 990. While board members are an important aspect of hospital organization, I do not include them in this analysis as they are beyond the scope of this paper. I extract text spanning 2009-2015, but I retain a much larger sample of hospitals by limiting to hospitals that have text information in 2010-2014. This yields 852 nonprofit hospitals that are matched to AHA data, and with complete text information in each year.

@@

    The hypothesis driven from the theory described in Section \ref{sec:forprofit} is that clinically trained executives put more weight on quality of care before the financial incentives on quality began. To investigate this hypothesis empirically, I merge all general for-profit hospitals in the AHA survey to my sample of nonprofit hospitals. Then, I compare readmission and mortality rates of always clinical executives and never clinical executives to for-profit hospitals after the incentive change. While I do not have information on leadership teams of for-profit hospitals, on average, they have revealed through their ownership status that they more heavily weigh profit. Thus, comparing different executive team behavior to for-profit behavior helps reveal how profit driven certain leadership teams are. Thus, I estimate the weighted synthetic difference-in-differences version of

>>>>

	\author{\vspace{10mm}Hanna Glenn\footnote{School of Economics, University of Queensland, hanna.glenn@uq.edu.au.} }

@@

    \begin{abstract}
		{\small
         Firm managers are often seen as key drivers of firm decisions, strategies, and performance. While extensive research links managerial characteristics to firm outcomes, less is known about how hospital executives influence responses to health care incentives. Health care policies in the US often focus on improving quality of care without increasing costs, yet it is unclear whether hospital leaders might contribute to this goal. Leveraging pay-for-performance incentives, I estimate the differential response to financial incentives on quality by different types of nonprofit hospital executives: those with and without a clinical training background. I find that hospitals without clinically trained executives respond more to the financial incentives than those with clinically trained executives. This can be explained by a model of hospital objectives in which clinical training affects hospital preferences towards patient welfare. These findings highlight the importance of considering leadership as an important factor in hospital decision making when adjusting incentives. 
		} 
	\end{abstract}

@@

    In corporate environments, executives are considered crucial drivers of a company's practices and performance, as they are the highest level of management that affects day-to-day operations of the organization. Executive heterogeneity has been linked to various actions and outcomes of firms; characteristics such as gender, age, and background are correlated with financial performance and behaviors such as employee diversity, wages, aggressiveness in mergers and acquisitions, and other corporate strategies.\footnote{\cite{bertrand2003managing}; \cite{matsa2011chipping}; \cite{matsa2011chipping}; \cite{adams2009women}; \cite{ahern2012changing}; \cite{flabbi2019female}; \cite{benmelech2015military}; \cite{custodio2013generalists}; \cite{frydman2019rising}} However, our understanding is typically limited to executives working for large, publicly traded corporations. Additionally, while correlations are an important first step to understanding the effects of leadership, little work has been done to uncover whether such correlations occur due to executive characteristics interacting with the objectives of firms.

    In this paper, I study the effect of specialized executive training on firm behavior when the firm experiences an external shock to their objective function. I do this specifically in the context of US hospitals for two reasons. First, in the US, health care makes up almost 20\% of spending (\cite{CMSNationalHealthExpenditure}), and a major goal of policymakers is to improve the quality of health care without increasing costs, as the US preforms poorly in these metrics compared to other similar countries. Yet, little is known about how leaders affect health care value in the hospitals they run. Second, nonprofit hospitals are particularly important in this context, and it is unclear whether existing findings on executive characteristics can be extrapolated to firms with presumably different incentives and objectives than those which have been previously studied.

    An executive characteristic relevant to this context is clinical training. Physicians can bring a unique combination of clinical and administration expertise to leadership, and they have insights that can potentially benefit their employees and improve quality (\cite{Stajduhar_2023}, \cite{Ahmed_2022}). However, hiring physician executives is controversial, as physicians are not always trained in business management, which can lead to poor practices and lower quality (\cite{HarvardBusinessReview2018}). Thus, I estimate how the presence of clinically trained executives on a hospital leadership team affects how the hospital makes decisions. I scrape executive names and titles of nonprofit hospitals in the US from publicly available Tax Form 990s. Combining this with other public hospital data, I construct a novel data set of nonprofit hospitals from 2010-2014 containing information on both executive team characteristics and quality metrics. While limited to nonprofits due to data availability, this sample of hospitals is an important driver of health care in the US. Private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up 36\% of hospitals and staff 107 beds on average (\cite{ASPE_2023}).

@@

    I use a theoretical model to investigate whether these results might be explained by differing preferences on the trade-off between profit and patient welfare, where I model these hospital objectives and allow the weight placed on each to vary with executive team type. Hospitals that place more weight on profit will naturally respond more drastically to a policy that adds quality directly into the profit function. Thus, a potential reason for the difference in response is that non-clinical executive teams are more profit driven. I investigate this empirically by comparing the response of hospitals with and without clinically trained executives to the response of for-profit hospitals, who have revealed that they are more profit driven by their ownership type. I find that nonprofit hospitals without clinically trained executives respond to the financial incentives on quality similarly to for-profits, while nonprofits without clinically trained executives respond differently than for-profits. For-profits and non-clinical executive teams change quality more in response to the incentive change than clinically trained executive teams.

    Finally, the question remains as to whether clinically trained executives are simply a signal of unchanging underlying hospital preferences (of, for example, the board of directors), or if they actively manage the hospital differently, effectively changing hospital objectives. I analyze this by leveraging changes in hospital propensity to hire clinical executives to decompose the full effect of clinical leadership into a signaling effect (clinical executives signal underlying hospital preferences), or managing effect (clinical executives manage the hospital differently). First, I show that hospital executive changes are not correlated with the change to financial incentives. To measure the signaling effect, I compare hospitals who ever have a clinically trained executive to hospitals who never have a clinically trained executive. For the managing effect, I limit to hospitals who ever have a clinically trained executive and compare those who have a clinically trained executive when the policies were announced to those who have clinically trained executives at other times. I find that the difference in response is driven by clinically trained executives managing the hospital differently.

@@

    \section{Theoretical Framework}\label{sec:theory}

@@

    \noindent where $\pi(\theta)$ is a profit function depending on quality $\theta$, $u(.)$ is extra utility gained from patient welfare, which is increasing and concave in quality $\theta$, and $\alpha\in[0,1]$ captures the weight placed on profit vs. patient welfare. There is an implicit stay-open condition throughout the model.

    I consider hospital choice of quality under pay-for-performance. In a world where performance is not incentivized through financial gain (as in traditional fee-for-service settings), hospital profit $\pi$ is maximized when the quantity of patients is maximized. However, pay-for-performance incentives incorporate quality into the profit function (\cite{dranove2011health}). Therefore, hospitals that only value financial profit will only consider quality if it is an input to the profit function, and therefore will respond more drastically to pay-for-performance than a hospital who values quality as an input to patient welfare before the incentives.

    I will now formalize this intuition mathematically. For simplicity, I abstract away from quality having a direct impact on quantity of patients or price.\footnote{As long as the relationship between quality and quantity/price are not correlated with $\alpha$, this does not change the conclusion of the model.} Taking the first order condition yields:

@@

    Now, consider incentives when the hospital is in a strictly fee-for-service environment. That is, quality benefits patients, but hospitals do not directly derive profit from having a high quality. Then, assuming there is a positive cost to increasing quality, $\pi(\theta)$ is decreasing in $\theta$. Then $\frac{du'(\theta)}{d\alpha} > 0$. Using the Implicit Function Theorem, I can write $\frac{d\theta}{d\alpha}$ as 
    $\frac{du'(\theta)/d\alpha}{du'(\theta)/d\theta}<0$, revealing that in a fee-for-service setting, the more weight the hospital places on profit, the lower quality of care chosen at baseline.

    Next, assume the hospital is in a pay-for-performance setting, where quality benefits patients and directly increases profit. Then, assuming the marginal financial benefit of increasing quality is greater than the marginal financial cost, $\pi(\theta)$ is increasing in $\theta$. Then $\frac{du'(\theta)}{d\alpha} < 0$, revealing that in a pay-for-performance setting, the more weight placed on profit, the higher quality of care chosen by the hospital. I combine the results found in each scenario into one response function that depends on $\alpha$, where $\theta_{p4p}$ is quality under pay-for-performance incentives and $\theta_{ffs}$ is quality under fee-for-service only. That is,

@@

    The first prediction of the model is that, all else held equal, hospitals that value profit will have a lower quality than those who value patient welfare in a fee-for-service setting. Second, a hospital's change in quality once pay-for-performance is enacted depends on how much weight the hospital places on profit vs. patient welfare. Particularly, hospitals with higher weight on profit respond more than hospitals with higher weight on patient welfare. Therefore, factors that affect weight $\alpha$ will be correlated with choice of quality, and with how quality changes under new pay-for-performance incentives. In this paper, I test the hypothesis that clinical experience on a hospital's leadership team is a factor that is correlated with weight $\alpha$.

@@

    For each hospital in the tax form data, I merge to hospital matches in the AHA data based on name and location.\footnote{I assess the observable differences between matched and non-matched hospitals in Appendix \ref{app:matched}, and conclude that the samples are overall similar apart from an under-sampling of hospitals that belong to systems.} For all hospitals that can be matched to a general hospital in the AHA data, I use optical character recognition (OCR) methods to extract the names, titles, and positions of each listed executive from its Tax Form 990. While board members are an important aspect of hospital organization, I do not include them in this analysis as they are beyond the scope of this paper. I extract text spanning 2009-2015, but I retain a much larger sample of hospitals by limiting to hospitals that have text information in 2010-2014. This yields 852 nonprofit hospitals that are matched to AHA data, and with complete text information in each year.

@@

    The hypothesis driven from the theory described in Section \ref{sec:theory} is that clinically trained executives put more weight on quality of care before the financial incentives on quality began. To investigate this hypothesis empirically, I merge all general for-profit hospitals in the AHA survey to my sample of nonprofit hospitals. Then, I compare readmission and mortality rates of always clinical executives and never clinical executives to for-profit hospitals after the incentive change. While I do not have information on leadership teams of for-profit hospitals, on average, they have revealed through their ownership status that they more heavily weigh profit. Thus, comparing different executive team behavior to for-profit behavior helps reveal how profit driven certain leadership teams are. Thus, I estimate the weighted synthetic difference-in-differences version of
